**Project Draft 1**

## Introduction

1.Life expectancy is an important indicator of a country's overall health and socioeconomic development. With Canada having an average of 82.2 years in 2024 (Statistics Canada, 2026), this provides residents with an indication of wellbeing based on solid statistics. Life expectancy is the average number of years a person is expected to live under current mortality conditions, which might be affected by different variables such as sex, income, and access to healthcare.

We will use the Life Expectancy (WHO) dataset from Kaggle, compiled from the WHO Global Health Observatory and UN economic data. It contains annual observations from 2000 to 2015 for 193 countries, covering health, demographic, socioeconomic, and immunization indicators (22 variables, 2938 observations).

Specifically, the data were collected from national health and economic reports by organizations such as the World Health Organization (WHO). These organizations compile, clean and standardize the data into databases, allowing indicators to be compared across different countries and years. Each observation represents a country in a specific year, with variables such as life expectancy, schooling, mortality rates, BMI, and development status.

The primary response variable we chose to analyze is life expectancy in years in 2015, serving as the dependent variable. Two explanatory variables are Schooling (number of years of schooling) and Status (country status, classified as developed or developing). We specifically chose these two because:\

a)  education, such as schooling, might be associated with opportunities, health knowledge, and quality of life, which can lead to better access to healthcare and living environments in the future.
b)  the status of a country is directly associated with healthcare facilities, access to clean resources, pollution, and economic factors such as income, which have a great impact on health.

Our study aims to learn more about how schooling is associated with life expectancy and whether life expectancy differs between developed and developing countries. We will conduct statistical summaries, linear regression, and visualizations to evaluate the significance of these relationships. To examine these relationships, a linear regression model will be used to evaluate how schooling and development status are associated with life expectancy.

The motivation for this study is to further understand, based on statistical analysis, whether these two factors have a significant impact on life expectancy across countries. By examining the relationship between schooling, development status, and life expectancy, the results may help identify areas where improvements can be made to better inform policy making and guide efforts aimed at improving public health and overall quality of life, ultimately helping people live longer and healthier lives.

## Statistical Data Analysis

| Variable              | Count | Median   | Mean     | SD       |
|-----------------------|-------|----------|----------|----------|
| Life expectancy (age) | 2928  | 72.10000 | 69.22493 | 9.523867 |
| Schooling (years)     | 2775  | 12.30000 | 11.99279 | 3.35892  |

Status (Categorical): Developed & Developing

The summary statistics show that the average life expectancy across countries is approximately 69.22 years, with a median of 72.10 years and a standard deviation of 9.52 years, indicating moderate variability among countries. The average years of schooling is approximately 11.99 years, with a median of 12.30 years and a standard deviation of 3.36 years, suggesting noticeable differences in education levels across countries.

------------------------------------------------------------------------

## Exploratory Data Analysis

We begin by exploring the relationship between schooling and life expectancy, as well as the role of development status.

### Life Expectancy by Development Status

```{r}
ggplot(data, aes(x = Status, y = LifeExpectancy, fill = Status)) +
  geom_boxplot() +
  labs(
    title = "Life Expectancy by Development Status",
    x = "Development Status",
    y = "Life Expectancy (years)"
  )
```

The boxplot suggests that developed countries tend to have higher life expectancy than developing countries. This indicates that development status may be an important predictor of life expectancy.

### Schooling vs Life Expectancy by Development Status

```{r}
data |>
  ggplot(aes(x = Schooling, y = LifeExpectancy, color = Status)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Life Expectancy vs Schooling by Development Status",
    x = "Schooling (years)",
    y = "Life Expectancy (years)"
  )
```

The scatter plot suggests a **positive relationship** between schooling and life expectancy: countries with higher average years of schooling tend to have higher life expectancy. In addition, developed countries generally appear to cluster in the upper-right portion of the plot, indicating both **higher schooling levels** and **higher life expectancy** compared with developing countries. This pattern suggests that both **Schooling** and **Status** are relevant predictors, and it also motivates considering an **interaction model** to determine whether the relationship between schooling and life expectancy differs by development status.

------------------------------------------------------------------------

## Model 1: Additive Linear Regression

We first fit an additive linear regression model including **Schooling** and **Status** as predictors:

$$
LifeExpectancy = \beta_0 + \beta_1(Schooling) + \beta_2(Status) + \epsilon
$$

### Model 1: Residuals vs Fitted Plot

```{r}
plot(model1$fitted.values, model1$residuals,
     main="Model 1: Residuals vs Fitted (Additive Model)",
     xlab="Fitted value",
     ylab="Residual")
abline(h=0)
```

The residuals vs fitted values plot shows that the residuals are generally scattered around zero without a clear systematic pattern. This suggests that the **linearity assumption** of the regression model is reasonably satisfied. In addition, the spread of the residuals appears relatively constant across the range of fitted values, indicating that the **homoscedasticity (constant variance) assumption** is reasonably met.

### Model 1: Normal Q–Q Plot

```{r}
qqnorm(resid(model1),
       main = "Model 1: Normal Q–Q Plot (Additive Model)")
qqline(resid(model1))
```

The Normal Q–Q plot shows that most residuals lie close to the reference line, suggesting that the residuals are **approximately normally distributed**. Although there are slight deviations in the tails, the overall pattern supports the assumption of normality for the regression residuals.

Overall, the diagnostic plots indicate that the additive regression model provides a **reasonable fit to the data**.

------------------------------------------------------------------------

## Model 2: Interaction Model

Next, we extend the regression model by including an **interaction term between Schooling and Status** to investigate whether the relationship between schooling and life expectancy differs between developed and developing countries.

$$
LifeExpectancy = \beta_0 + \beta_1(Schooling) + \beta_2(Status) + \beta_3(Schooling \times Status) + \epsilon
$$

### Model 2: Residuals vs Fitted Plot

```{r}
plot(model2$fitted.values, model2$residuals,
     main = "Model 2: Residuals vs Fitted (Interaction Model)",
     xlab = "Fitted Values",
     ylab = "Residuals")
abline(h = 0)
```

The residuals vs fitted values plot for the interaction model also shows residuals randomly scattered around zero. There is no clear systematic structure or strong curvature, suggesting that the **linearity assumption remains reasonable**. The variability of the residuals also appears fairly constant across fitted values, indicating that the **constant variance assumption is satisfied**.

### Model 2: Normal Q–Q Plot

```{r}
qqnorm(resid(model2),
       main = "Model 2: Normal Q–Q Plot (Interaction Model)")
qqline(resid(model2))
```

The Q–Q plot shows that the residuals largely follow the reference line, with minor deviations in the extreme tails. This indicates that the **normality assumption for the residuals is approximately satisfied**.

Overall, the diagnostic plots suggest that the **interaction model also provides an appropriate fit**, and including the interaction term allows us to explore whether the effect of schooling on life expectancy differs by development status.

### Model 1: Regression Results

The estimated regression model is:

LifeExpectancy = beta0 + beta1\* Schooling + beta2 \* Status

The results form the summary table show that:

-   **Schooling** has a coefficient of 2.098 with p-value, 2e-16, which shows that with increasing of one year of schooling, life expectancy increases by approximately **2.098 years**, holding the other variable constant. This effect is statistically significant.

-   **Status** has a coefficient of -1.826 with p-value equals to 0.104, suggesting that developing countries have lower life expectancy than developed countries on average. However, this effect is **not statistically significant at the 0.05 level**.

The model has **R²** with the value of **0.6745**, indicating that schooling and development status are able to explain approximately **67.45%** of the variation in life expectancy.

### Model 2: Interaction Model Results

The estimated model is:

LifeExpectancy = beta0 + beta1\* Schooling + beta2 \* Status + beta3 \* Schooling \* Status

The results show that:

-   The interaction term, **Schooling × StatusDeveloping** has a coefficient of 1.479 with the p-value of 0.006, which is statistically significant. This indicates that the effect of schooling on life expectancy **differs between developed and developing countries**.

-   In **developed countries**, which is the baseline group, the effect of schooling is 0.730 with p-value of 0.156, which is not statistically significant at the significance level of 0.05.

-   In **developing countries**, the effect of schooling becomes:

    0.730 + 1.479 = **2.209**

This indicates that for each additional year of schooling, life expectancy increases by approximately **2.21 years** in developing countries, showing a stronger association compared to developed countries.

-   **Schooling** has a coefficient of 0.7303 with p-value, 0.15584, which shows that with increasing of one year of schooling, life expectancy increases by approximately **0.7303 years**, holding the other variable constant, but it is not statistically significant at the significance level of 0.05.

-   The coefficient for StatusDeveloping is -25.805 with the p-value of 0.003 indicates that developing countries have substantially lower life expectancy than developed countries, when keeping the other variables constant.

The model has **R²** with the value of **0.6887**, indicating that schooling, the interaction term and development status are able to explain approximately **68.87%** of the variation in life expectancy.

According to the value of the **adj R²**, the model with interaction term has slightly larger value, 0.6832, than the additive model, with the value of the **adj R²**, 0.6707. This suggests that the model with interaction may provide a better fit to the data than the additive model.

### Variable Selection

```{r}
#forward selection
forward_model<-regsubsets(LifeExpectancy ~ ., data=data_sel, method="forward")
ss = summary(forward_model)
metrics = data.frame(
  R2 = ss$rsq,
  AdjR2 = ss$adjr2,
  Cp = ss$cp
)
metrics
summary(forward_model)$which[3, ]
```

```{r}
#backward selection
backward_model <- regsubsets(LifeExpectancy ~ ., data=data_sel, method="backward")

ss2=summary(backward_model)
metrics_backward = data.frame(
  R2 = ss2$rsq,
  AdjR2 = ss2$adjr2,
  Cp = ss2$cp
)
metrics_backward
summary(backward_model)$which[3, ]
```

From both the forward and backward selection results, three predictor model was selected because its Mallows’ value (Cp 4.629796) is closest to the number of predictors (k = 3), while the adjusted R\^2 (0.8854601 at k = 3) shows slight changes with additional predictors.

The best three predictors identified for both models are the same: Adult Mortality, Hepatitis B, Income composition of resources.

### Conclusion

### Statement of AI Use

### Appendix
