---
title: "The Association Between Schooling and Life Expectancy: A Comparative Analysis of Developed and Developing Countries"
author:
  - name: "Group Members"
    affiliation: "STAT 306, University of British Columbia"
date: "March 24, 2025"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup, include=FALSE}
library(tidyverse)
library(leaps)
library(car)
library(knitr)

# Read and prepare data
data <- read_csv("https://drive.google.com/uc?export=download&id=1A_Nkqsxh4ymFIJDj7Fbo1SOCoVujhbjb")

data <- data %>%
  rename(LifeExpectancy = `Life expectancy`) %>%
  filter(Year == 2015)

data_sel <- data %>%
  select(-Year, -Country, -Alcohol, -`Total expenditure`) %>%
  na.omit()

# Convert Status to factor
data_sel$Status <- as.factor(data_sel$Status)

# Fit models
model1 <- lm(LifeExpectancy ~ `Income composition of resources` +
              `Adult Mortality` + `Hepatitis B` + Schooling + Status,
            data = data_sel)

model2 <- lm(LifeExpectancy ~ `Income composition of resources` +
              `Adult Mortality` + `Hepatitis B` + Schooling * Status,
            data = data_sel)

# Subset selection
forward_model <- regsubsets(LifeExpectancy ~ ., data = data_sel, method = "forward")
ss <- summary(forward_model)

backward_model <- regsubsets(LifeExpectancy ~ ., data = data_sel, method = "backward")
ss2 <- summary(backward_model)

exhaustive_model <- regsubsets(LifeExpectancy ~ ., data = data_sel, method = "exhaustive")
ss3 <- summary(exhaustive_model)
```

# Introduction

Life expectancy is a fundamental indicator of a country's overall health and socioeconomic development. In 2024, Canada's average life expectancy reached 82.2 years, reflecting the nation's strong healthcare system and high quality of life. However, life expectancy varies considerably across countries, ranging from under 50 years in some nations to over 80 years in others. Understanding the factors that contribute to these differences is important for informing public health policy and development strategies.

This study investigates the relationship between education, as measured by average years of schooling, and life expectancy across countries. Education may influence health outcomes through multiple pathways, including improved health literacy, better employment opportunities, higher income, and greater access to healthcare resources. Additionally, we examine whether the association between education and life expectancy differs between developed and developing countries.

## Data Source and Description

The data for this study are sourced from the World Health Organization (WHO) Global Health Observatory and United Nations economic databases, compiled into the *Life Expectancy (WHO)* dataset available on Kaggle. The dataset contains annual observations from 2000 to 2015 for 193 countries, with variables related to health outcomes, demographic characteristics, socioeconomic indicators, and immunization coverage.

For this analysis, we focus on data from **2015**, the most recent year available in the dataset. After removing observations with missing values, the final dataset consists of **130 countries** with complete information on all variables used in the model.

### Variables of Interest

| Variable                                      | Type         | Description                                                                              |
|------------------------|------------------------|------------------------|
| **LifeExpectancy** (Response)                 | Quantitative | Average number of years a newborn is expected to live under current mortality conditions |
| **Schooling** (Primary predictor)             | Quantitative | Average number of years of schooling among adults                                        |
| **Status** (Primary predictor)                | Categorical  | Country development classification: Developed or Developing                              |
| **Adult Mortality** (Control)                 | Quantitative | Probability of dying between ages 15 and 60 per 1000 population                          |
| **Hepatitis B** (Control)                     | Quantitative | Percentage of one-year-olds immunized against Hepatitis B                                |
| **Income composition of resources** (Control) | Quantitative | Index measuring the distribution of income and resources                                 |

Adult Mortality was included because it reflects deaths in the working-age population, which serves as an indicator for national life expectancy.

Hepatitis B was included to indicate the strength of the healthcare system, which reduces the infectious disease rate and improves survival.

Income Composition of Resources was included because it reflects national socioeconomic conditions that influence access to healthcare, living quality, and nutrients, which are critical for impacting life expectancy.

### Data Cleaning

The following data cleaning steps were performed:

1.  Filtered observations to retain only **2015** data.
2.  Removed variables not used in the current analysis (`Year`, `Country`, `Alcohol`, and `Total expenditure`).
3.  Excluded observations with missing values using listwise deletion.
4.  Converted `Status` to a factor for categorical analysis.

# Exploratory Data Analysis

In this section, we will perform exploratory data analysis (EDA) to explore the variable and their relationships with life expectancy. This step helps us to recognize and identify patterns, distributions, and potential issues, and provides initial quantitative evidence about which predictors might be related to life expectancy.

## Summary Statistics

```{r summary-stats}
summary_stats <- data_sel %>%
  summarise(
    `Life Expectancy (years)` = paste0(round(mean(LifeExpectancy), 2), " (", round(sd(LifeExpectancy), 2), ")"),
    `Schooling (years)` = paste0(round(mean(Schooling), 2), " (", round(sd(Schooling), 2), ")"),
    `Adult Mortality` = paste0(round(mean(`Adult Mortality`), 2), " (", round(sd(`Adult Mortality`), 2), ")"),
    `Hepatitis B (%)` = paste0(round(mean(`Hepatitis B`), 2), " (", round(sd(`Hepatitis B`), 2), ")"),
    `Income Composition` = paste0(round(mean(`Income composition of resources`), 2), " (", round(sd(`Income composition of resources`), 2), ")")
  ) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Mean (SD)")

kable(summary_stats, caption = "Summary Statistics for Key Variables (N = 130)")
```

The average life expectancy across countries is about **71.6 years** (SD = 8.1), and the average years of schooling is about **12.5 years** (SD = 3.1), indicating substantial variation across countries.

## Life Expectancy by Development Status

```{r boxplot-status}
ggplot(data_sel, aes(x = Status, y = LifeExpectancy, fill = Status)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Life Expectancy by Development Status (2015)",
    x = "Development Status",
    y = "Life Expectancy (years)"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("steelblue", "orange"))
```

Developed countries generally have higher life expectancy than developing countries, although there is overlap between the two groups.

## Schooling and Life Expectancy

```{r scatter-plot}
ggplot(data_sel, aes(x = Schooling, y = LifeExpectancy, color = Status)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Life Expectancy vs. Schooling by Development Status (2015)",
    x = "Average Years of Schooling",
    y = "Life Expectancy (years)",
    color = "Development Status"
  ) +
  theme_minimal() +
  scale_color_manual(values = c("steelblue", "orange"))
```

The scatter plot suggests a positive relationship between schooling and life expectancy. The fitted lines appear to have similar slopes for developed and developing countries, which motivates checking whether an interaction term is needed.

# Statistical Analysis

## Variable Selection

Before fitting the main regression models, subset selection was used to identify a reasonable set of control variables. We focused on Mallows' $C_p$, which prefers models whose $C_p$ values are close to the number of parameters $p = k + 1$ (where $k$ is the number of predictors).

```{r subset-selection}
cp_table <- data.frame(
  `Number of Variables (k)` = 1:length(ss3$cp),
  `Cp` = round(ss3$cp, 2),
  `p = k + 1` = 2:(length(ss3$cp) + 1)
)

kable(cp_table, caption = "Mallows' Cp Values for Exhaustive Subset Selection")
```

The **3-variable model** has a $C_p$ value of **4.63**, which is closest to $p = 4$, indicating it is the optimal model according to Mallows' $C_p$ criterion. The selected control variables are:

-   **Adult Mortality**: a key indicator of population health\
-   **Hepatitis B**: represents immunization coverage and healthcare access\
-   **Income composition of resources**: captures socioeconomic conditions

These three variables were retained as controls in all subsequent regression models.

## Model 1: Additive Linear Regression

We first fit an additive linear regression model that includes the three control variables along with our predictors of interest:

$$
\begin{aligned}
\text{LifeExpectancy} ={}& \beta_0 + \beta_1(\text{Adult Mortality}) + \beta_2(\text{Hepatitis B}) \\
&+ \beta_3(\text{Income composition}) + \beta_4(\text{Schooling}) + \beta_5(\text{Status}) + \epsilon
\end{aligned}
$$

```{r model1-results}
summary(model1)
```

```{r model1-diagnostics, fig.height=4, fig.width=8}
par(mfrow = c(1, 2))
plot(model1$fitted.values, model1$residuals,
     main = "Model 1: Residuals vs Fitted",
     xlab = "Fitted Values", ylab = "Residuals",
     pch = 19, cex = 0.5)
abline(h = 0, col = "red", lty = 2)

qqnorm(resid(model1), main = "Model 1: Normal Q-Q Plot",
       pch = 19, cex = 0.5)
qqline(resid(model1), col = "red", lty = 2)
par(mfrow = c(1, 1))
```

### Model 1 Results

**Hypothesis Testing**

For each coefficient, we test:

-   $H_0: \beta_j = 0$ (no effect on life expectancy)
-   $H_1: \beta_j \neq 0$ (significant effect)

| Variable           | Estimate | Std. Error | t-value | p-value  | Decision                 |
|------------|------------|------------|------------|------------|------------|
| (Intercept)        | 47.381   | 2.206      | 21.483  | \< 0.001 | Reject $H_0$             |
| Income composition | 36.807   | 4.611      | 7.983   | \< 0.001 | Reject $H_0$             |
| Adult Mortality    | -0.025   | 0.003      | -8.217  | \< 0.001 | Reject $H_0$             |
| Hepatitis B        | 0.046    | 0.010      | 4.532   | \< 0.001 | Reject $H_0$             |
| Schooling          | -0.081   | 0.233      | -0.349  | 0.727    | **Fail to reject** $H_0$ |
| StatusDeveloping   | -0.144   | 0.810      | -0.177  | 0.860    | **Fail to reject** $H_0$ |

**Overall Model Fit**

-   Multiple R-squared: **0.8882**
-   Adjusted R-squared: **0.8837**
-   F-statistic: 197.1 on 5 and 124 DF, p-value: \< 2.2e-16

The significant F-test (p \< 0.001) indicates that the model as a whole explains a substantial portion of the variation in life expectancy. The non-significant coefficients for Schooling (p = 0.727) and Status (p = 0.860) suggest that, after controlling for adult mortality, Hepatitis B vaccination, and income composition, these variables do not have statistically significant independent effects on life expectancy.

## Model 2: Interaction Model

We then tested whether the effect of schooling on life expectancy differs between developed and developing countries by adding an interaction term:

$$
\begin{aligned}
\text{LifeExpectancy} ={}& \beta_0 + \beta_1(\text{Adult Mortality}) + \beta_2(\text{Hepatitis B}) + \beta_3(\text{Income composition}) \\
&+ \beta_4(\text{Schooling}) + \beta_5(\text{Status}) + \beta_6(\text{Schooling} \times \text{Status}) + \epsilon
\end{aligned}
$$

```{r model2-results}
summary(model2)
```

```{r model2-diagnostics, fig.height=4, fig.width=8}
par(mfrow = c(1, 2))
plot(model2$fitted.values, model2$residuals,
     main = "Model 2: Residuals vs Fitted",
     xlab = "Fitted Values", ylab = "Residuals",
     pch = 19, cex = 0.5)
abline(h = 0, col = "red", lty = 2)

qqnorm(resid(model2), main = "Model 2: Normal Q-Q Plot",
       pch = 19, cex = 0.5)
qqline(resid(model2), col = "red", lty = 2)
par(mfrow = c(1, 1))
```

### Model 2 Results

**Hypothesis Testing for Interaction**

To answer whether the association between schooling and life expectancy differs by development status, we test:

-   $H_0: \beta_6 = 0$ (no difference between developed and developing countries)
-   $H_1: \beta_6 \neq 0$ (the effect differs)

| Variable                       | Estimate | Std. Error | t-value | p-value   | Decision                 |
|------------|------------|------------|------------|------------|------------|
| (Intercept)                    | 54.006   | 7.296      | 7.402   | \< 0.001  | Reject $H_0$             |
| Income composition             | 35.210   | 4.908      | 7.174   | \< 0.001  | Reject $H_0$             |
| Adult Mortality                | -0.026   | 0.003      | -8.268  | \< 0.001  | Reject $H_0$             |
| Hepatitis B                    | 0.046    | 0.010      | 4.535   | \< 0.001  | Reject $H_0$             |
| Schooling                      | -0.401   | 0.409      | -0.982  | 0.328     | Fail to reject $H_0$     |
| StatusDeveloping               | -6.928   | 7.167      | -0.967  | 0.336     | Fail to reject $H_0$     |
| **Schooling:StatusDeveloping** | 0.421    | 0.442      | 0.953   | **0.343** | **Fail to reject** $H_0$ |

**Interpretation of the Interaction**

The interaction term has a p-value of **0.343**, which is greater than 0.05. Therefore, we fail to reject the null hypothesis. There is **no sufficient evidence** that the effect of schooling on life expectancy differs between developed and developing countries.

-   For **developed countries** (baseline): effect of schooling = -0.401 (p = 0.328, not significant)
-   For **developing countries**: effect of schooling = -0.401 + 0.421 = 0.020 (not significant)

**Overall Model Fit**

-   Multiple R-squared: **0.8891**
-   Adjusted R-squared: **0.8837**
-   Residual standard error: 2.728 on 123 degrees of freedom

## Model Selection

We compare the additive model (Model 1) and the interaction model (Model 2) to determine which provides a more appropriate description of the data.

The interaction term (Schooling × Status) is **not statistically significant** (p = 0.343). This indicates that adding the interaction term does not significantly improve the model's ability to explain variation in life expectancy. Additionally, the adjusted R² values are nearly identical for both models (0.8837 for both), suggesting that the interaction term does not provide additional explanatory power.

Therefore, we select the **additive model (Model 1)** as our final model. This choice is supported by the principle of parsimony: when a simpler model explains the data nearly as well as a more complex model, the simpler model is preferred.

## Multicollinearity Assessment

Variance Inflation Factors (VIFs) were used to assess multicollinearity in the additive model.

```{r vif}
vif_table <- data.frame(
  Variable = names(vif(model1)),
  VIF = round(vif(model1), 2)
)

kable(vif_table, caption = "Variance Inflation Factor (VIF) for Additive Model")
```

All VIF values are below 10, suggesting that multicollinearity is not a serious concern. The moderate correlation between Schooling (VIF = 7.19) and Income composition (VIF = 8.43) is expected, as education and income are naturally related, but does not invalidate the model.

# Discussion

## Addressing the Research Question

This study asked two questions:

1.  **How is schooling associated with life expectancy?**
2.  **Does this association differ between developed and developing countries?**

Based on our analysis, we find:

1.  **Schooling does not have a statistically significant independent effect on life expectancy** after controlling for adult mortality, Hepatitis B vaccination, and income composition (p = 0.727). The positive bivariate relationship observed in the scatter plot is largely explained by these confounding factors.

2.  **The association between schooling and life expectancy does not differ between developed and developing countries**. The interaction term (Schooling × Status) was not statistically significant (p = 0.343), indicating that the relationship between education and life expectancy is consistent across development contexts.

3.  **The strongest predictors of life expectancy in the final model are:**

    -   **Income composition of resources** (positive, p \< 0.001)
    -   **Adult Mortality** (negative, p \< 0.001)
    -   **Hepatitis B vaccination** (positive, p \< 0.001)

These results suggest that education's apparent relationship with life expectancy is **indirect**, operating through improvements in health outcomes and economic conditions. When these mediating factors are accounted for, education itself does not have a direct effect on longevity.

### Policy Implications

The findings suggest that policies aimed at improving life expectancy should prioritize: - **Direct health interventions**: Reducing adult mortality and increasing vaccination coverage - **Economic development**: Improving income distribution and economic conditions

While education remains important for individual and societal development, its effect on population health appears to operate through these more direct channels.

## Limitations

Several limitations should be noted:

1.  **Cross-sectional design**: The analysis uses only 2015 data, which provides a snapshot rather than capturing changes over time. This prevents causal inference about the direction of relationships.

2.  **Country-level aggregation**: Relationships observed at the country level may not reflect individual-level relationships (ecological fallacy). Individual-level data would be needed to confirm these findings.

3.  **Measurement limitations**: Years of schooling is an imperfect measure of educational quality. Additionally, the binary classification of developed/developing may mask substantial heterogeneity within each group.

4.  **Potential omitted variables**: Other factors such as healthcare expenditure, environmental quality, political stability, and cultural factors were not included in the analysis.

# Conclusion

This study examined the relationship between education and life expectancy using 2015 data from 130 countries. After controlling for key health and socioeconomic factors (adult mortality, Hepatitis B vaccination, and income composition), we found:

1.  **Education does not have a significant independent effect on life expectancy**. The apparent bivariate relationship is explained by confounding factors such as adult mortality, vaccination coverage, and income distribution.

2.  **The association between education and life expectancy does not differ between developed and developing countries**. The interaction term was not statistically significant (p = 0.343), indicating that the relationship is consistent across development contexts.

3.  **The strongest predictors of life expectancy** are income composition (positive), adult mortality (negative), and Hepatitis B vaccination coverage (positive), collectively explaining 88.8% of the variation in life expectancy.

These findings suggest that policies aimed at improving life expectancy should prioritize direct health interventions and economic development. Education remains valuable, but its impact on population health appears to be indirect, mediated through improvements in health outcomes and economic conditions.

# Statement of AI Use

We used generative AI (ChatGPT) for the following purposes:

1.  **Code debugging**: Assistance with troubleshooting R code errors and optimizing visualizations.
2.  **Writing assistance**: Help with structuring the report, improving clarity, and correcting grammatical errors.
3.  **Interpretation guidance**: Discussion of statistical concepts and interpretation of regression results.

All prompts used during the writing process are provided in the Appendix.

# References

Link, B. G., & Phelan, J. (1995). Social conditions as fundamental causes of disease. *Journal of Health and Social Behavior*, 80-94.

Statistics Canada. (2026). *Life expectancy at birth, by sex, Canada*.

World Health Organization. (2020). *Global Health Observatory data repository*.

# Appendix

## AI Prompts Used

1.  "How do I interpret the VIF output in R for a model with an interaction term?"
2.  "What is the difference between additive and interaction models in linear regression?"
3.  "How should I explain the Mallows' Cp criterion in my report?"
4.  "Help me structure the discussion section for a regression analysis on life expectancy."
5.  "Can you help me create a clean table of regression coefficients in R Markdown?"
