---
title: "Project Draft 1"
author:
  - name: "Ruiting Jiang, Mark Cao, Lillian Chen, Jiayi Lyu"
    affiliation: "STAT 306, University of British Columbia"
date: "March 23, 2025"
format:
  pdf:
    documentclass: article
    pdf-engine: pdflatex
    fontsize: 10pt
    geometry: margin=1in
    number-sections: true
    toc: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup, include=FALSE}
library(tidyverse)
library(leaps)
library(car)
library(knitr)
library(broom)

options(knitr.kable.NA = "")

# Read and clean data
data <- read_csv("https://drive.google.com/uc?export=download&id=1A_Nkqsxh4ymFIJDj7Fbo1SOCoVujhbjb")

data <- data %>%
  rename(LifeExpectancy = `Life expectancy`) %>%
  filter(Year == 2015)

data_sel <- data %>%
  select(-Year, -Country, -Alcohol, -`Total expenditure`) %>%
  na.omit()

data_sel$Status <- as.factor(data_sel$Status)

# Preliminary models
model1 <- lm(LifeExpectancy ~ Schooling + Status, data = data_sel)
model2 <- lm(LifeExpectancy ~ Schooling * Status, data = data_sel)

# Variable selection
forward_model  <- regsubsets(LifeExpectancy ~ ., data = data_sel, method = "forward")
backward_model <- regsubsets(LifeExpectancy ~ ., data = data_sel, method = "backward")
best_model     <- regsubsets(LifeExpectancy ~ ., data = data_sel, method = "exhaustive")

ss_e <- summary(best_model)

# Final models with selected controls
model3 <- lm(LifeExpectancy ~ `Income composition of resources` +
               `Adult Mortality` + `Hepatitis B` + Schooling + Status,
             data = data_sel)

model4 <- lm(LifeExpectancy ~ `Income composition of resources` +
               `Adult Mortality` + `Hepatitis B` + Schooling * Status,
             data = data_sel)

# Tables
sample_summary <- tibble(
  Statistic = c("Sample size", "Mean life expectancy", "Median life expectancy",
                "SD of life expectancy", "Mean schooling", "SD of schooling"),
  Value = c(
    nrow(data_sel),
    round(mean(data_sel$LifeExpectancy), 2),
    round(median(data_sel$LifeExpectancy), 2),
    round(sd(data_sel$LifeExpectancy), 2),
    round(mean(data_sel$Schooling), 2),
    round(sd(data_sel$Schooling), 2)
  )
)

cp_table <- tibble(
  `Number of predictors` = 1:length(ss_e$cp),
  `Adjusted R^2` = round(ss_e$adjr2, 4),
  `Cp` = round(ss_e$cp, 4),
  `Target p = k + 1` = 2:(length(ss_e$cp) + 1)
)

coef_m1 <- tidy(model1) %>% mutate(across(where(is.numeric), ~ round(.x, 4)))
coef_m2 <- tidy(model2) %>% mutate(across(where(is.numeric), ~ round(.x, 4)))
coef_m3 <- tidy(model3) %>% mutate(across(where(is.numeric), ~ round(.x, 4)))
coef_m4 <- tidy(model4) %>% mutate(across(where(is.numeric), ~ round(.x, 4)))

vif_m3 <- tibble(
  Variable = names(vif(model3)),
  VIF = round(as.numeric(vif(model3)), 3)
)
```

# Introduction

Life expectancy is an important indicator of a country’s overall health and socioeconomic development. For example, Canada had an average life expectancy of 82.2 years in 2024 (Statistics Canada, 2026), reflecting a high standard of living and a well-developed healthcare system. In general, life expectancy represents the average number of years a person is expected to live under current mortality conditions, and it may be influenced by a range of factors such as income, access to healthcare, and overall living conditions.

In this study, we use the *Life Expectancy (WHO)* dataset from Kaggle, which compiles country-level health and socioeconomic indicators from the World Health Organization (WHO) and related United Nations sources. The full dataset contains annual observations from 2000 to 2015 for 193 countries and includes variables on mortality, immunization, income, education, and other demographic and health characteristics.

The data are collected from national health and economic reports and are compiled, cleaned, and standardized by organizations such as the WHO to allow meaningful comparisons across countries and over time. Each observation represents a country in a specific year and includes variables such as life expectancy, schooling, mortality rates, BMI, and development status.

For this analysis, we focus on a **cross-sectional dataset using observations from the year 2015**, the most recent year available. Restricting the data to a single year allows us to avoid complications arising from repeated observations of the same country over time and makes the standard regression assumptions more reasonable.

The primary response variable in this analysis is life expectancy in years for 2015, which serves as the dependent variable. Life expectancy varies over time and may be influenced by external factors not captured in the dataset, such as global health events. For example, events such as the SARS outbreak in 2003 and the H1N1 (Swine Flu) pandemic in 2009 could temporarily affect life expectancy and introduce variability unrelated to the variables of interest.

To reduce this source of variation and maintain a consistent cross-sectional framework, we restrict our analysis to the year 2015, which we treat as a relatively stable reference point.

The two primary explanatory variables are **Schooling** (average years of schooling) and **Status** (country classification as developed or developing).

Our primary research question is:

> **How is schooling associated with life expectancy, and does this association differ between developed and developing countries?**

We focus on **Schooling** and **Status** because education may influence health knowledge, quality of life, and access to opportunities, while development status reflects broader differences in healthcare systems, infrastructure, and living conditions. By examining these relationships using statistical summaries, visualizations, and linear regression models, we aim to better understand which factors are most strongly associated with life expectancy across countries and how these relationships vary across different development contexts.

# Data Source and Description

Each observation in the dataset represents a country in a given year. To maintain a cross-sectional framework and avoid dependence due to repeated measurements over time, we restrict the analysis to observations from **2015**.

We remove variables not used in the current analysis (`Year`, `Country`, `Alcohol`, and `Total expenditure`) and apply listwise deletion to retain complete cases. The final dataset consists of **130 countries**.

We removed `Country` from the analysis because it contains a large number of unique categories, which would introduce a high number of parameters if included directly in a regression model. Instead, we use the variable `Status`, as a meaningful grouping of countries, which allows us to capture broad structural differences while maintaining interpretability.

The variables `Alcohol` and `Total expenditure` were removed due to a substantial number of missing values. We assume that these missing values occur without a systematic pattern and therefore do not introduce significant bias after listwise deletion. However, the missingness is not random, this could be a limitation of the analysis.

## Variables Used in the Analysis

While the dataset contains many variables, we focus on a subset that is most relevant to our research question.

| Variable | Type | Description |
|------------------------|------------------------|------------------------|
| **LifeExpectancy** | Quantitative | Average number of years a newborn is expected to live under current mortality conditions |
| **Schooling** | Quantitative | Average number of years of schooling among adults |
| **Status** | Categorical | Country development classification: Developed or Developing |
| **Adult Mortality** | Quantitative | Probability of dying between ages 15 and 60 per 1000 population |
| **Hepatitis B** | Quantitative | Percentage of one-year-olds immunized against Hepatitis B |
| **Income composition of resources** | Quantitative | Index measuring the distribution of income and resources |

## Data Cleaning

The following data cleaning steps were performed:

1.  Filtered observations to retain only **2015** data.
2.  Removed `Country`, `Alcohol`, and `Total expenditure`.
3.  Excluded observations with missing values using listwise deletion.
4.  Converted `Status` to a factor for categorical analysis.

## Statistical Summary

```{r}
kable(sample_summary, caption = "Summary statistics for the analysis sample")
```

The average life expectancy in the analysis sample is **71.62 years**, with a median of **73.90** and a standard deviation of **8.12**. Average schooling is **12.54 years**, indicating noticeable cross-country variation in educational attainment.

# Analysis

## Exploratory Data Analysis

We begin by examining how life expectancy varies by development status and schooling.

```{r fig.height=3.5, fig.width=5.2}
ggplot(data_sel, aes(x = Status, y = LifeExpectancy, fill = Status)) +
  geom_boxplot() +
  labs(
    title = "Life Expectancy by Development Status",
    x = "Development Status",
    y = "Life Expectancy (years)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

The boxplot suggests that developed countries tend to have higher life expectancy than developing countries, although there is overlap between the two groups. After a closer examination, we pinpointed the source of this overlap. As of 2015, the category of "Developing countries" contain both countries that struggle with basic health insurance and now "developed countries" such as Canada, making the variation in life expectancy in devloping countries high and creating the overlap between the two statues.

```{r fig.height=3.5, fig.width=5.4}
ggplot(data_sel, aes(x = Schooling, y = LifeExpectancy, color = Status)) +
  geom_point(alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Life Expectancy vs Schooling by Development Status",
    x = "Schooling (years)",
    y = "Life Expectancy (years)"
  ) +
  theme_minimal()
```

The scatter plot suggests a positive relationship between schooling and life expectancy. It also suggests that the slope for developing countries may be steeper than the slope for developed countries, which motivates considering an interaction model in addition to an additive model. Also, there appears to be slightly more variation in life expectancy for smaller values of schooling, which we will do a distribution check to address and from there, potentially consider data transformation.

## Distribution check

```{r fig.height=3.5, fig.width=5.4}
# Distribution of key variables
ggplot(data_sel, aes(x = LifeExpectancy)) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of Life Expectancy")

ggplot(data_sel, aes(x = Schooling)) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of Schooling")

ggplot(data_sel, aes(x = `Adult Mortality`)) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of Adult Mortality")

ggplot(data_sel, aes(x = `Income composition of resources`)) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of Income Composition")
```

We examine the distributions of key variables to assess whether transformations may be necessary. Most variables appear reasonably symmetric, including Schooling, although Adult Mortality shows some right-skewness. This suggests that a transformation such as a logarithmic transformation may be considered.

## Log transformation check

```{r fig.height=3.5, fig.width=5.4}
model_log <- lm(LifeExpectancy ~ log(`Adult Mortality`) +
                  `Income composition of resources` +
                  `Hepatitis B` + Schooling + Status,
                data = data_sel)

summary(model_log)

# Compare models
AIC(model3, model_log)
```

We also explored potential transformations for skewed variables. In particular, we applied a log transformation to Adult Mortality to account for its right-skewed distribution.

The transformed model did not lead to a substantial improvement in model fit or interpretability, as measured by AIC and overall conclusions. Therefore, for simplicity and ease of interpretation, we proceed with the original (untransformed) variables in the final model.

This suggests that the linear model assumptions are reasonably satisfied without requiring major transformations.

## Preliminary Regression Analysis

We now move from exploratory data analysis to statistical modelling. The previous section focuses on visualizing patterns and relationships in the data; this section uses linear regression and models to evaluate the relationship between schooling and life expectancy. In this section, we will fit preliminary regression models to understand the relationship between schooling and life expectancy and to explore whether this relationship is impacted by development status. We first consider an additive model, followed by an interaction model.

We first fit two simple models using only the variables directly named in the research question.

### Model 1: Additive Model

$$
\text{LifeExpectancy} = \beta_0 + \beta_1(\text{Schooling}) + \beta_2(\text{Status}) + \epsilon
$$ For the schooling effect in Model 1, we test:

$$
H_0:\beta_1 = 0 \qquad \text{vs} \qquad H_A:\beta_1 \ne 0
$$

For the status effect in Model 1, we test:

$$
H_0:\beta_2 = 0 \qquad \text{vs} \qquad H_A:\beta_2 \ne 0
$$

```{r}
kable(coef_m1, caption = "Model 1 coefficient estimates")
```

Model 1 shows a strong positive association between schooling and life expectancy. The coefficient for **Schooling** is **2.2260** with a p-value less than 0.001, suggesting that, holding status constant, an additional year of schooling is associated with an increase of about **2.23 years** in life expectancy. The coefficient for **StatusDeveloping** is negative but not statistically significant at the 0.05 level.

Model 1 has an (R\^2) of **0.6533** and an adjusted (R\^2) of **0.6478**.

### Model 2: Interaction Model

$$
\text{LifeExpectancy} = \beta_0 + \beta_1(\text{Schooling}) + \beta_2(\text{Status}) + \beta_3(\text{Schooling} \times \text{Status}) + \epsilon
$$ For the interaction term in Model 2, we test:

$$
H_0:\beta_3 = 0 \qquad \text{vs} \qquad H_A:\beta_3 \ne 0
$$

```{r}
kable(coef_m2, caption = "Model 2 coefficient estimates")
```

In Model 2, the interaction term **Schooling × StatusDeveloping** is statistically significant (p = **0.0082**), suggesting that in this simple model the effect of schooling appears to differ by development status. For developed countries, the schooling slope is **0.4622**, while for developing countries the slope is **0.4622 + 1.8892 = 2.3514**. Model 2 has a slightly larger adjusted (R\^2) (**0.6643**) than Model 1.

These preliminary results suggest a possible interaction, but they do not account for additional confounding variables. We therefore next perform variable selection and refit the models with selected controls.

## Variable Selection

To identify a reasonable set of control variables, we applied forward selection, backward selection, and exhaustive best subset selection using `regsubsets()`. We focus on Mallows’ (C_p) together with adjusted (R\^2). In the exhaustive search results below, the three-predictor model has (C_p = 4.6298), which is closest to the target value (p = k+1 = 4), while adjusted (R\^2) changes only modestly as additional predictors are added.

```{r}
kable(cp_table, caption = "Exhaustive subset selection results")
```

In addition to examining Mallows' (C_p) and adjusted (R\^2), we also inspected which variables were selected for each model size. The table above shows the variables included in each exhaustive subset model.

In particular, the best model, being the three-predictor model includes:

-   **Adult Mortality**
-   **Hepatitis B**
-   **Income composition of resources**

This model has a (C_p) value closest to the target value (p = k+1), indicating a good balance between model fit and model complexity. These three variables were also selected consistently across the forward, backward, and exhaustive procedures.

We therefore retain these variables as control variables in the final stage of the analysis in order to better isolate the relationship between our primary variables of interest, **Schooling** and **Status**, and life expectancy.

## Final Regression Analysis

### Model 3: Additive Model With Controls

$$
\begin{aligned}
\text{LifeExpectancy} = &\beta_0 + \beta_1(\text{Income composition}) + \beta_2(\text{Adult Mortality}) \\
&+ \beta_3(\text{Hepatitis B}) + \beta_4(\text{Schooling}) + \beta_5(\text{Status}) + \epsilon
\end{aligned}
$$

For the schooling and status effects in Model 3, we test:

$$
H_0:\beta_4 = 0 \qquad \text{vs} \qquad H_A:\beta_4 \ne 0
$$

$$
H_0:\beta_5 = 0 \qquad \text{vs} \qquad H_A:\beta_5 \ne 0
$$

```{r}
kable(coef_m3, caption = "Model 3 coefficient estimates")
```

IThe results from Model 3 indicate that the selected control variables are all statistically significant and have meaningful associations with life expectancy.

-   **Income composition of resources** has a strong positive coefficient (**36.81**, p \< 0.001). This suggests that, holding all other variables constant, an increase in this index is associated with a substantial increase in life expectancy. Since this variable captures broader socioeconomic conditions, this result highlights the importance of economic development and resource distribution in determining population health.

-   **Adult Mortality** has a statistically significant negative coefficient (**-0.0252**, p \< 0.001). This indicates that higher adult mortality rates are associated with lower life expectancy, which is consistent with expectations, as this variable directly reflects population health risks.

-   **Hepatitis B immunization** has a positive and statistically significant coefficient (**0.0458**, p \< 0.001). This suggests that higher immunization coverage is associated with increased life expectancy, reflecting the role of preventive healthcare and public health interventions.

However, after controlling for these variables:

-   **Schooling** is no longer statistically significant (p = **0.727**). We fail to reject the null hypothesis $H_0: \beta_{\text{Schooling}} = 0$, indicating that there is no evidence of an independent association between schooling and life expectancy once these control variables are included.

-   **Status** is also not statistically significant (p = **0.860**). We fail to reject the null hypothesis $H_0: \beta_{\text{Status}} = 0$, suggesting that, conditional on the included controls, there is no statistically significant difference in life expectancy between developed and developing countries.

This change compared to the preliminary models suggests that the apparent relationship between schooling (and development status) and life expectancy may be explained by confounding variables such as economic conditions, mortality rates, and immunization coverage. In other words, schooling may be indirectly related to life expectancy through these factors rather than having a direct effect.

The model has an (R\^2) of **0.8882** and an adjusted (R\^2) of **0.8837**, indicating that approximately **88.8%** of the variation in life expectancy is explained by the variables included in the model. This high explanatory power suggests that the selected variables capture key determinants of life expectancy at the country level.

### Model 4: Interaction Model With Controls

We specifically examined the interaction between educational achievement and development level, as our research question directly addresses whether the relationship between educational achievement and life expectancy differs between developed and developing countries. We did not include interaction terms involving control variables, as doing so would significantly increase the model’s complexity, reduce its interpretability, and stray from the core issues of primary concern. Our objective is to determine whether the impact of educational achievement varies across different levels of development.

$$
\begin{aligned}
\text{LifeExpectancy} = &\beta_0 + \beta_1(\text{Income composition}) + \beta_2(\text{Adult Mortality}) + \beta_3(\text{Hepatitis B}) \\
&+ \beta_4(\text{Schooling}) + \beta_5(\text{Status}) + \beta_6(\text{Schooling} \times \text{Status}) + \epsilon
\end{aligned}
$$

For the interaction term in Model 4, we test:

$$
H_0:\beta_6 = 0 \qquad \text{vs} \qquad H_A:\beta_6 \ne 0
$$

```{r}
kable(coef_m4, caption = "Model 4 coefficient estimates")
```

In Model 4, the interaction term **Schooling × StatusDeveloping** is not statistically significant (p = **0.343**). The adjusted (R\^2) is **0.8837**, essentially identical to Model 3. Thus, after controlling for adult mortality, Hepatitis B immunization, and income composition, there is no evidence that the relationship between schooling and life expectancy differs between developed and developing countries.

Because the interaction is not significant and the adjusted (R\^2) does not improve, we select **Model 3** as the final model by the principle of parsimony.

```{r}
AIC(model1, model2, model3, model4)
```

We further compare the models using the Akaike Information Criterion (AIC), which balances model fit and complexity.

Model 3 has the lowest AIC (637.60) among all models, indicating the best trade-off between goodness of fit and model simplicity. Although Model 4 includes an interaction term, it has a slightly higher AIC (638.64), suggesting that the added complexity is not justified.

Therefore, both the hypothesis test for the interaction term and the AIC comparison support selecting Model 3 as the final model.

## Diagnostics and Multicollinearity

```{r fig.height=3.5, fig.width=6.8}
par(mfrow = c(1, 2))
plot(model3$fitted.values, model3$residuals,
     main = "Model 3: Residuals vs Fitted",
     xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)

qqnorm(resid(model3), main = "Model 3: Normal Q-Q Plot")
qqline(resid(model3), col = "red", lty = 2)
par(mfrow = c(1, 1))
```

The residuals vs fitted plot shows no strong systematic pattern, suggesting that the linearity and constant variance assumptions are reasonably satisfied. The Q-Q plot indicates that the residuals are approximately normal, with only modest deviations in the tails.These issues are not severe but should be acknowledged as limitations when interpreting the model.

We assessed multicollinearity using variance inflation factors (VIFs) for the final additive model.

```{r}
kable(vif_m3, caption = "VIF values for the final model")
```

The largest VIF values are for **Income composition of resources** (**8.433**) and **Schooling** (**7.188**), indicating moderate collinearity but not an extreme level for this analysis. The remaining VIF values are low. Overall, multicollinearity does not appear severe enough to invalidate the final model, although it may contribute to reduced significance for schooling after the controls are added.

# Conclusion

In this study, we applied statistical modeling and analysis to explore how schooling and development status are associated with life expectancy using the WHO Life Expectancy dataset. Initial simple regression models suggested that schooling had a positive relationship with life expectancy, and this relationship appeared stronger in developing countries than in developed countries. However, after variable selection identified key control variables—adult mortality, Hepatitis B vaccination, and income composition of resources—the effects of both schooling and development status were no longer statistically significant in the full additive model. The interaction term was also not significant in the controlled interaction model (p = 0.343).

Instead, income composition of resources and Hepatitis B immunization showed strong positive associations with life expectancy, while adult mortality showed a strong negative association. These results suggest that life expectancy is influenced more directly by economic and health-related conditions rather than by schooling and development status alone, reflecting the combined impact of broader social, medical, and national factors.

In conclusion, schooling and development status initially appeared to be associated with life expectancy, but the final regression model showed that their effects were not statistically significant after accounting for other factors. This result indicates that life expectancy is more strongly related to variables such as income composition of resources, adult mortality, and Hepatitis B immunization.

Two limitations should be noted. First, the study is based on observational country-level data from a single year, so the results should be interpreted as associations rather than causal effects. Second, other relevant determinants of life expectancy may be omitted from the final model.

## Submission Details

The final report (PDF) was submitted via Gradescope, and the R/qmd code along with the data files were submitted via Canvas by **Jiayi Lyu** on behalf of the group.

# Statement of AI Use

Generative AI was used to assist with language revision, grammar checking, and formatting of the report. All statistical analyses, model selections, and interpretations were performed by the group members and verified against the course material.

# Appendix

## Distribution comparison: original vs log-transformed

```{r fig.height=3.5, fig.width=6.8}
par(mfrow = c(1,2))
hist(data_sel$`Adult Mortality`, main = "Original Adult Mortality", xlab = "Adult Mortality")
hist(log(data_sel$`Adult Mortality`), main = "Log-transformed Adult Mortality", xlab = "log(Adult Mortality)")
par(mfrow = c(1,1))
```

## AI Prompts Used

1.  "Help me improve the grammar and clarity of this paragraph."
2.  "Can you help me format the regression coefficient table for a report?"

All AI-generated suggestions were reviewed and edited by the group before inclusion in the final report.
